---
title: "Summary and Visualization Functions"
author: "multivar package"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Summary and Visualization Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

The `multivar` package provides three main functions for summarizing and visualizing fitted models:

1. **`summary_multivar()`** - Comprehensive model summary
2. **`print_dynamics()`** - Visualize dynamics matrices
3. **`print_edge_prevalence()`** - Show edge prevalence across subjects (K>1)

This vignette demonstrates how to use these functions to explore and interpret your multivar model results.

---

# Getting Started

First, let's load the package and fit a simple model:

```{r load-package}
library(multivar)
```

```{r fit-simple-model, results='hide', message=FALSE}
# Simulate data with K=2 subjects
set.seed(123)
sim <- multivar_sim(
  k = 2,
  d = 4,
  n = 100,
  prop_fill_com = 0.3,
  prop_fill_ind = 0.15,
  sigma = diag(4),
  lb = 0.1,
  ub = 0.9
)

# Fit the model
object <- constructModel(sim$data, intercept = FALSE)
fit <- cv.multivar(object)
```

---

# Model Summary: `summary_multivar()`

The `summary_multivar()` function provides a comprehensive overview of your fitted model.

## Basic Usage

```{r summary-basic}
summary_multivar(fit)
```

## What Information is Displayed?

### Dataset Information
- **Number of datasets (subjects)**: The K value
- **Number of timepoints**: Handles varying lengths across subjects
- **Number of variables**: The d value

### Model Specification
- **Time-varying parameters**: Whether the model includes TVP
- **Period ranges**: For TVP models, shows time ranges for each period
- **Intercepts**: Whether intercepts are included and penalized

### Adaptive Weights
- **Method**: LASSO, RIDGE, etc.
- **Type**: Adaptive or Standard
- **Lambda selection**: Minimum CV error or 1 SE rule
- **Cross-validation**: Type and number of folds

### Overview of Dynamics
- **Edge counts**: Number of non-zero entries
- **Sparsity**: Percentage of zero entries
- **Value ranges**: Range of non-zero coefficients

---

# Visualizing Dynamics: `print_dynamics()`

The `print_dynamics()` function displays dynamics matrices with clear formatting and time subscripts.

## Direct Use with Fit Objects

One of the key features is that you can pass the fit object directly:

```{r print-dynamics-fit}
print_dynamics(fit, digits = 3)
```

## Key Features

- **Time subscripts**: Rows labeled as `V1_t`, columns as `V1_{t-1}`
- **Dots for zeros**: Uses `.` to display zero entries for better readability
- **Summary statistics**: Shows sparsity and value ranges
- **Automatic routing**: Detects TVP models and formats accordingly

## Customization Options

```{r print-dynamics-custom, eval=FALSE}
# More decimal places
print_dynamics(fit, digits = 4)

# Show zeros as "0" instead of "."
print_dynamics(fit, zero_char = "0")

# Without time subscripts
print_dynamics(fit, time_labels = FALSE)
```

---

# Time-Varying Models

Let's demonstrate the summary functions with a time-varying parameter (TVP) model.

## Fitting a TVP Model

```{r fit-tvp-model, results='hide', message=FALSE}
# Create time-varying dynamics
phi_period1 <- matrix(c(
  0.3,  0.4,  0.0,  0.0,
  0.0,  0.5,  0.3,  0.0,
  0.4,  0.0,  0.2,  0.0,
  0.0,  0.3,  0.0,  0.4
), nrow = 4, byrow = TRUE)

phi_period2 <- matrix(c(
  0.4,  0.3,  0.0,  0.2,
  0.0,  0.6,  0.4,  0.0,
  0.3,  0.0,  0.3,  0.0,
  0.0,  0.4,  0.0,  0.5
), nrow = 4, byrow = TRUE)

phi_period3 <- matrix(c(
  0.5,  0.2,  0.0,  0.0,
  0.0,  0.4,  0.5,  0.0,
  0.5,  0.0,  0.4,  0.2,
  0.0,  0.5,  0.0,  0.3
), nrow = 4, byrow = TRUE)

# Generate data
phi_list <- c(
  rep(list(phi_period1), 80),
  rep(list(phi_period2), 80),
  rep(list(phi_period3), 80)
)

data1 <- multivar:::var_sim_growth(
  n = 240,
  phi = phi_list,
  sigma = diag(4),
  intercept = rep(0, 4)
)

# Fit TVP model
object_tvp <- constructModel(
  list(data1),
  intercept = FALSE,
  tvp = TRUE,
  breaks = list(c(80, 160)),
  cv = "blocked",
  nfolds = 5
)

fit_tvp <- cv.multivar(object_tvp)
```

## TVP Model Summary

```{r summary-tvp}
summary_multivar(fit_tvp)
```

Notice how the summary automatically detects the TVP structure and shows:
- Number of periods
- Period ranges (t=1-79, t=80-159, t=160-239)
- Per-period edge statistics

## TVP Dynamics with Custom Labels

```{r print-dynamics-tvp}
print_dynamics(
  fit_tvp,
  period_labels = c("Early", "Middle", "Late"),
  digits = 3
)
```

---

# Edge Prevalence Across Subjects

For models with multiple subjects (K>1), `print_edge_prevalence()` shows which edges are present across subjects.

## Proportion Display

Shows what proportion of subjects have each edge:

```{r edge-prevalence-proportion}
print_edge_prevalence(fit, type = "proportion", digits = 2)
```

## Count Display

Shows the actual count of subjects with each edge:

```{r edge-prevalence-count}
print_edge_prevalence(fit, type = "count")
```

## Interpreting the Output

The summary statistics tell you:

- **Universal edges (100%)**: Present in all subjects
- **Common edges (>50%)**: Present in more than half of subjects
- **Rare edges (<50%)**: Present in less than half of subjects
- **Absent edges (0%)**: Not present in any subject

This helps identify:
- **Shared network structure**: Universal and common edges represent the common dynamics
- **Individual variation**: Rare edges show subject-specific patterns
- **Network sparsity**: Absent edges indicate where no relationships exist

---

# Comparison with K=3 Model

Let's see how edge prevalence looks with more subjects:

```{r fit-k3-model, results='hide', message=FALSE}
# Simulate K=3 model
sim3 <- multivar_sim(
  k = 3,
  d = 4,
  n = 100,
  prop_fill_com = 0.3,
  prop_fill_ind = 0.15,
  sigma = diag(4),
  lb = 0.1,
  ub = 0.9
)

object3 <- constructModel(sim3$data, intercept = FALSE)
fit3 <- cv.multivar(object3)
```

```{r edge-prevalence-k3}
print_edge_prevalence(fit3, type = "proportion", digits = 2)
```

With more subjects, you can see more nuanced patterns of edge prevalence.

---

# Complete Workflow Example

Here's a typical workflow combining all three functions:

```{r complete-workflow, eval=FALSE}
# 1. Fit your model
fit <- cv.multivar(object)

# 2. Get comprehensive summary
summary_multivar(fit)

# 3. Visualize dynamics
print_dynamics(fit)

# 4. For K>1: Check edge prevalence
print_edge_prevalence(fit, type = "proportion")
```

---

# Tips and Best Practices

## When to Use Each Function

- **`summary_multivar()`**: Always use this first to understand your model structure
- **`print_dynamics()`**: Use to examine specific dynamics matrices with clear formatting
- **`print_edge_prevalence()`**: Use for K>1 models to understand shared vs. unique structure

## Customization

All functions support customization:

```{r customization-examples, eval=FALSE}
# Summary (no customization needed - automatic)
summary_multivar(fit)

# Dynamics with custom formatting
print_dynamics(
  fit,
  digits = 4,              # More precision
  zero_char = "—",         # Custom zero character
  period_labels = c(...)   # For TVP models
)

# Edge prevalence with custom display
print_edge_prevalence(
  fit,
  type = "count",          # Show counts instead of proportions
  digits = 3,              # Decimal places for proportions
  zero_char = "—"          # Custom zero character
)
```

## Working with Different Model Types

### K=1 Models
- `summary_multivar()` shows single-subject statistics
- `print_dynamics()` displays one matrix
- `print_edge_prevalence()` returns a message (not applicable)

### K>1 Models
- `summary_multivar()` shows common + subject-specific statistics
- `print_dynamics()` displays matrices for all subjects
- `print_edge_prevalence()` shows edge overlap across subjects

### TVP Models
- `summary_multivar()` shows per-period statistics
- `print_dynamics()` displays time-varying matrices with period labels
- `print_edge_prevalence()` shows prevalence separately for each period (if K>1)


