---
title: "Summary and Visualization Functions in multivar"
author: "Zachary F. Fisher"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: pdflatex
    keep_tex: false
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{xcolor}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[L]{Summary Functions}
  - \fancyhead[R]{\texttt{multivar} Package}
  - \usepackage{geometry}
  - \geometry{margin=1in}
geometry: margin=1in
fontsize: 11pt
vignette: >
  %\VignetteIndexEntry{Summary and Visualization Functions (PDF)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  out.width = "85%"
)
```

\newpage

# Introduction

The `multivar` package provides three main functions for summarizing and visualizing fitted models:

1. **`summary_multivar()`** -- Comprehensive model summary with dataset information, model specifications, and dynamics overview
2. **`print_dynamics()`** -- Visualize dynamics matrices with clear formatting and time subscripts
3. **`print_edge_prevalence()`** -- Display edge prevalence across subjects for multi-subject models ($K>1$)

This vignette demonstrates how to use these functions to explore and interpret multivar model results effectively.

\medskip

**Key Features:**

- Automatic detection of model types (TVP vs. non-TVP, $K=1$ vs. $K>1$)
- Clean, publication-ready formatting
- Flexible customization options
- Seamless integration with fitted model objects

---

# Getting Started

First, load the package and fit a simple model to demonstrate the summary functions.

## Loading the Package

```{r load-package}
library(multivar)
```

## Simulating and Fitting a Model

We'll simulate data with $K=2$ subjects, $d=4$ variables, and $n=100$ timepoints:

```{r fit-simple-model, results='hide', message=FALSE}
# Simulate data with K=2 subjects
set.seed(123)
sim <- multivar_sim(
  k = 2,
  d = 4,
  n = 100,
  prop_fill_com = 0.3,
  prop_fill_ind = 0.15,
  sigma = diag(4),
  lb = 0.1,
  ub = 0.9
)

# Fit the model
object <- constructModel(sim$data, intercept = FALSE)
fit <- cv.multivar(object)
```

\newpage

# Model Summary: `summary_multivar()`

The `summary_multivar()` function provides a comprehensive overview of your fitted model, including dataset characteristics, model specifications, and an overview of the estimated dynamics.

## Basic Usage

```{r summary-basic}
summary_multivar(fit)
```

## Information Displayed

The summary includes four main sections:

### Dataset Information

- **Number of datasets (subjects)**: The $K$ value
- **Number of timepoints**: Handles both equal and varying lengths across subjects
- **Number of variables**: The $d$ value (dimension of the VAR system)

### Model Specification

- **Time-varying parameters (TVP)**: Whether the model includes time-varying dynamics
- **Period ranges**: For TVP models, shows time ranges for each period
- **Intercepts**: Whether intercepts are included and how they are penalized

### Adaptive Weights

- **Method**: LASSO, RIDGE, or other penalization methods
- **Type**: Adaptive or Standard weighting
- **Lambda selection**: Minimum CV error (`lambda.min`) or 1 SE rule (`lambda.1se`)
- **Cross-validation**: Type (blocked or rolling) and number of folds

### Overview of Dynamics

- **Edge counts**: Number of non-zero entries in transition matrices
- **Sparsity**: Percentage of zero entries
- **Value ranges**: Range of non-zero coefficient values
- **Per-subject breakdown**: For $K>1$ models, shows both common and subject-specific statistics

\newpage

# Visualizing Dynamics: `print_dynamics()`

The `print_dynamics()` function displays dynamics matrices with clear formatting, time subscripts, and summary statistics. A key feature is that you can pass fit objects directly.

## Direct Use with Fit Objects

```{r print-dynamics-fit}
print_dynamics(fit, digits = 3)
```

## Key Features

- **Time subscripts**: Rows labeled as $V_i(t)$, columns as $V_j(t-1)$
- **Dots for zeros**: Uses `.` to display zero entries for better readability
- **Alignment**: Positive values include leading space to align with negative values
- **Summary statistics**: Shows sparsity and value ranges
- **Automatic routing**: Detects TVP models and formats accordingly

## Customization Options

```{r print-dynamics-custom, eval=FALSE}
# More decimal places
print_dynamics(fit, digits = 4)

# Show zeros as "0" instead of "."
print_dynamics(fit, zero_char = "0")

# Without time subscripts
print_dynamics(fit, time_labels = FALSE)
```

\newpage

# Time-Varying Parameter (TVP) Models

The summary functions automatically detect and handle time-varying parameter models. Let's demonstrate with a TVP example.

## Fitting a TVP Model

```{r fit-tvp-model, results='hide', message=FALSE}
# Create time-varying dynamics
phi_period1 <- matrix(c(
  0.3,  0.4,  0.0,  0.0,
  0.0,  0.5,  0.3,  0.0,
  0.4,  0.0,  0.2,  0.0,
  0.0,  0.3,  0.0,  0.4
), nrow = 4, byrow = TRUE)

phi_period2 <- matrix(c(
  0.4,  0.3,  0.0,  0.2,
  0.0,  0.6,  0.4,  0.0,
  0.3,  0.0,  0.3,  0.0,
  0.0,  0.4,  0.0,  0.5
), nrow = 4, byrow = TRUE)

phi_period3 <- matrix(c(
  0.5,  0.2,  0.0,  0.0,
  0.0,  0.4,  0.5,  0.0,
  0.5,  0.0,  0.4,  0.2,
  0.0,  0.5,  0.0,  0.3
), nrow = 4, byrow = TRUE)

# Generate data
phi_list <- c(
  rep(list(phi_period1), 80),
  rep(list(phi_period2), 80),
  rep(list(phi_period3), 80)
)

data1 <- multivar:::var_sim_growth(
  n = 240,
  phi = phi_list,
  sigma = diag(4),
  intercept = rep(0, 4)
)

# Fit TVP model
object_tvp <- constructModel(
  list(data1),
  intercept = FALSE,
  tvp = TRUE,
  breaks = list(c(80, 160)),
  cv = "blocked",
  nfolds = 5
)

fit_tvp <- cv.multivar(object_tvp)
```

## TVP Model Summary

```{r summary-tvp}
summary_multivar(fit_tvp)
```

Notice how the summary automatically detects the TVP structure and displays:

- Number of periods (3 in this case)
- Period ranges: $t=1$--79, $t=80$--159, $t=160$--239
- Per-period edge statistics showing how network structure evolves over time

## TVP Dynamics with Custom Labels

```{r print-dynamics-tvp}
print_dynamics(
  fit_tvp,
  period_labels = c("Early", "Middle", "Late"),
  digits = 3
)
```

The period labels provide interpretable names for each time window, making it easier to communicate results.

\newpage

# Edge Prevalence Across Subjects

For models with multiple subjects ($K>1$), `print_edge_prevalence()` shows which edges are present across subjects. This helps identify universal dynamics versus subject-specific patterns.

## Proportion Display

Shows what proportion of subjects have each edge:

```{r edge-prevalence-proportion}
print_edge_prevalence(fit, type = "proportion", digits = 2)
```

## Count Display

Shows the actual count of subjects with each edge:

```{r edge-prevalence-count}
print_edge_prevalence(fit, type = "count")
```

## Interpreting the Output

The summary statistics identify:

- **Universal edges (100\%)**: Present in all subjects -- these represent shared dynamics
- **Common edges ($>50\%$)**: Present in more than half of subjects
- **Rare edges ($<50\%$)**: Present in less than half of subjects -- potentially subject-specific
- **Absent edges (0\%)**: Not present in any subject

This breakdown helps researchers understand:

1. **Shared network structure**: Universal and common edges
2. **Individual variation**: Rare edges showing subject-specific patterns
3. **Network sparsity**: Absent edges indicate where no relationships exist

\newpage

# Comparison with $K=3$ Model

Edge prevalence patterns become more nuanced with additional subjects:

```{r fit-k3-model, results='hide', message=FALSE}
# Simulate K=3 model
sim3 <- multivar_sim(
  k = 3,
  d = 4,
  n = 100,
  prop_fill_com = 0.3,
  prop_fill_ind = 0.15,
  sigma = diag(4),
  lb = 0.1,
  ub = 0.9
)

object3 <- constructModel(sim3$data, intercept = FALSE)
fit3 <- cv.multivar(object3)
```

```{r edge-prevalence-k3}
print_edge_prevalence(fit3, type = "proportion", digits = 2)
```

With more subjects, you can observe more granular patterns of edge prevalence, helping to distinguish between highly consistent edges, moderately consistent edges, and idiosyncratic edges.

\newpage

# Complete Workflow Example

A typical workflow combining all three functions:

```{r complete-workflow, eval=FALSE}
# 1. Fit your model
fit <- cv.multivar(object)

# 2. Get comprehensive summary
summary_multivar(fit)

# 3. Visualize dynamics
print_dynamics(fit)

# 4. For K>1: Check edge prevalence
print_edge_prevalence(fit, type = "proportion")
```

---

# Tips and Best Practices

## When to Use Each Function

- **`summary_multivar()`**: Always use this first to understand model structure and get an overview
- **`print_dynamics()`**: Use to examine specific dynamics matrices with publication-ready formatting
- **`print_edge_prevalence()`**: Use for $K>1$ models to understand shared vs. unique network structure

## Customization Examples

All functions support flexible customization:

```{r customization-examples, eval=FALSE}
# Summary (automatic - no customization needed)
summary_multivar(fit)

# Dynamics with custom formatting
print_dynamics(
  fit,
  digits = 4,              # More precision
  zero_char = "—",         # Custom zero character
  period_labels = c(...)   # For TVP models
)

# Edge prevalence with custom display
print_edge_prevalence(
  fit,
  type = "count",          # Show counts instead of proportions
  digits = 3,              # Decimal places for proportions
  zero_char = "—"          # Custom zero character
)
```

## Working with Different Model Types

### $K=1$ Models (Single Subject)

- `summary_multivar()` shows single-subject statistics
- `print_dynamics()` displays one matrix
- `print_edge_prevalence()` returns a message (not applicable for single subject)

### $K>1$ Models (Multiple Subjects)

- `summary_multivar()` shows common + subject-specific statistics
- `print_dynamics()` displays matrices for all subjects
- `print_edge_prevalence()` shows edge overlap across subjects

### TVP Models (Time-Varying)

- `summary_multivar()` shows per-period statistics
- `print_dynamics()` displays time-varying matrices with period labels
- `print_edge_prevalence()` shows prevalence separately for each period (if $K>1$)

\newpage

# Summary

The multivar summary functions provide:

\begin{itemize}
\item \textbf{Comprehensive information} about model structure and results
\item \textbf{Clear visualization} of dynamics with readable, publication-ready formatting
\item \textbf{Cross-subject comparisons} via edge prevalence analysis
\item \textbf{Automatic detection} of model types (TVP, $K=1$ vs. $K>1$)
\item \textbf{Flexible customization} for different display and reporting needs
\end{itemize}

These tools make it easy to explore, interpret, and communicate multivar model results in both exploratory data analysis and formal research reporting.

---

# References

Fisher, Z. F., Kim, Y., Fredrickson, B. L., & Pipiras, V. (2022). Penalized estimation of multiple-subject Vector Autoregressive models. *Psychometrika*, *87*(1), 1--28.

